Characterizing dynamical transitions by statistical complexity measures based on ordinal pattern transition networks

### 2021-02-12 manuscript_Huang_rev.tex
Reik's final changes.

The situation becomes even more challenging when our modified SCMs are considered, since they are based on transition frequencies instead of pattern frequencies. For $D!$ different ordinal patterns for embedding dimension $D$, there may exist $D!^2$ possible source--destination pairs. However, if we consider $\tau=1$ and transitions corresponding to shifting the time index by one time step only, it is clear that from the $D$ components of the source permutation, $D-1$ also contribute to the corresponding destination pattern without changing their mutual order relationships \cite{BorgesAMC2019,PessaPRE2019}. Hence, the new state contributing to the destination pattern can only lead to at most $D$ different ordinal patterns, i.e., we have at most $D\cdot D!$ different pattern transitions. However, for other cases like $\tau>1$ and/or larger time shifts between ordinal patterns to be considered, up to $D!^2$ different pairs of patterns and, hence, transitions could occur in the worst case. In general, we conclude that the far higher number of pattern transitions as compared to patterns themselves puts severe restrictions to the choice of $D$ in practical applications of the new SCMs in the context of time series of a given length $L$.
}

@@

\subsection{\textcolor{red}{Effect} of embedding parameters} \label{sec:embeddings}
There are two important algorithmic parameters in constructing OPTNs -- the embedding parameters $D$ and $\tau$. The choice of $D$ and $\tau$ can markedly affect the results for the associated SCMs since $D$ and $\tau$ should compromise the essential time scales of the dynamics in the underlying time series. In common applications of ordinal pattern analysis (e.g., the estimation of the permutation entropy $\mathcal{S}_O$), the choice of optimal (or just pragmatic) values of $D$ and $\tau$ depends on the field of application as systematically summarized in Ref.~\cite{Riedl2013}. {\color{red}Throughout the remainder of this work, we will consider $D$ as a free parameter, which is varied in the range of $D \in \{2,\ldots,7\}$, although for the sake of brevity, we will mainly show the results for $D = 6$ as illustration. The associated choice of the embedding delay $\tau$ is commonly less crucial in nonlinear time series analysis methods. In our numerical examples discussed below, the corresponding selection will be mainly guided by the attempt to highlight the fact that lower entropy values are expected for periodic as compared to chaotic behavior. In order to cope with a single parameter only, we will particularly use $\tau=D$ for the artificial (deterministic and stochastic) model systems considered in the following, independent of the particular value of $D$. We emphasize that this choice is not based on specific considerations of performance optimization, but simply reflects convenience and serves solely illustrative purposes. Two examples for results obtained with $D=6$ and $\tau=1$ for two well-known model systems can be found in the Supplementary Material Sec.~SM2, Figs.~S4 and S5.}

@@

\textcolor{red}{More importantly, for sufficiently large $D$ the resulting complexity--entropy pairs} essentially cover exclusively the left uprising branch of the curve for the SCM \textcolor{red}{accounting for} transition frequencies (Figs.~\ref{fig:CElogistic}(b, e) and \ref{fig:CElogistic}(c, f)), which distinctively differs from the original permutation entropy based SCM. \textcolor{red}{In this regard, one may note that this branch implies an almost linear dependency between $\mathcal{C}_W$ and $\mathcal{H}_{W}$ (Fig.~\ref{fig:CElogistic}(e)), respectively, $\mathcal{C}_{E}$ and $\mathcal{H}_{E}$ (Fig.~\ref{fig:CElogistic}(f)), which especially applies for small entropy values within an increasingly wide range as $D$ increases. This linear increase of the maximum complexity versus entropy curve for sufficiently large embedding dimension is expected from theoretical considerations~\cite{SMCbook2010}. Some more discussions have been provided in the Supplementary Material (Sec. SM3, Fig. S6-S8). }

@@

{\color{red}The behavior of the different SCMs relative to their respective lower and upper bounds can be further highlighted by considering normalized quantities. Specifically, for a given $\mathcal{H}$ and $D$, we define a normalized complexity $\mathcal{C}^{N} = (\mathcal{C} - \mathcal{C}_{min}) / (\mathcal{C}_{max} - \mathcal{C}_{min})$, which is theoretically bound to the interval $[0,1]$. Large values of those normalized complexities close to 1 indicate a good alignment with the maximum complexity curve. The corresponding results for the logistic map are shown in Fig.~\ref{fig:CHlogCmax}. As expected, for the traditional $\mathcal{C}_{O}$, the behavior is characterized by many points in the right half plane (large entropies, see Fig.~\ref{fig:CHlogCmax}(a)) for all considered embedding dimensions. In contrast, there are no points in the range of $\mathcal{H} >0.8$ for $\mathcal{C}_W$ and $\mathcal{C}_E$ (Fig.~\ref{fig:CHlogCmax}(b, c)). Note that in Fig.~\ref{fig:CHlogCmax}(c), some normalized complexity values larger than 1 arise at small entropies due to numerical artifacts when the control parameter $r$ is close to a bifurcation point, which results in long transients (as will be explained in Sec.~\ref{sec:transi}).}

@@

Especially for the new OPTN based SCMs, we find a partial overlap between the entropy values of fBm and the logistic map, while the associated SCM values are however clearly distinct except for the \textcolor{red}{fBm realizations} with the lowest entropy \textcolor{red}{values}, which correspond to the most nonstationary situations ($H$ close to 1). 
\textcolor{red}{In general, we observe for all three complexity--entropy planes of fBm that with increasing $D$, the associated range of entropy values extends further towards the left (lower entropies). Taken both, upward and leftward shift of the complexity--entropy curves individually may therefore enhance the risk of confusing deterministic-chaotic and stochastic processes in the complexity--entropy plane even with our alternative SCMs. However, our results indicate that} combining information on the position in the complexity--entropy plane and on the stationarity could allow a clear distinction between chaotic and stochastic dynamics even in extreme situations \textcolor{red}{(e.g., a large value of $H$)}. Notably, for the stationary fGn, all \textcolor{red}{three complexity--entropy planes} are confined to very high entropy and low complexity values -- a range where no combinations have been found for the new transition frequency based SCMs for the logistic map.

@@

As a general observation, we find that all three SCMs clearly follow the changes in the bifurcation diagram, for instance, showing distinct values in periodic windows (highlighted by grey color in Fig.~\ref{fig:bifurcation} \textcolor{red}{and Supplementary Material Fig.~S4, where the obtained results are shown for the same embedding dimension $D=6$, but two different values of the embedding delay $\tau$}). Other than $\mathcal{C}_O$, the two transition frequency based SCMs behave in a similar way as the Lyapunov exponent. Specifically, as the control parameter $r$ is increased, the chaoticity level grows gradually, reaching a maximum at $r = 4$, which is reflected by an increase in statistical complexity. A similar increase is absent in the pattern frequency based complexity measure $\mathcal{C}_O$ (Fig.~\ref{fig:bifurcation}(c)). In this context, we note that it is an established fact that pattern frequency based SCMs often do not trace the growth in the chaoticity of the logistic map~\cite{MartinPLA2003}, which can however be improved by using Wooter's distance function instead of the Shannon-Jensen divergence employed in this work. From this point of view, the pattern transition frequency based SCMs $\mathcal{C}_W$ and $\mathcal{C}_E$ exhibit some more informative behavior in the sense that they track the growth of the level of chaoticity.

@@

While we have focused so far exclusively on the case of the time-discrete logistic map, it may be interesting to study whether a similar behavior can also be obtained for time-continuous deterministic dynamical systems. To this end, we will illustrate a corresponding analysis for the example of the chaotic R\"ossler system \cite{Roessler1976} 
\begin{eqnarray}
\dot{x} &=& -y-z, \nonumber \\
\dot{y} &=& x+0.2y, \\
\dot{z} &=& b+z(x-5.7) \nonumber
\end{eqnarray}
while varying the control parameter $b$. From a conceptual perspective, constructing OPTNs for time series from continuous systems faces certain additional practical challenges, including the proper choice of sampling frequency and embedding parameters, which depend on the particular time scales of the system. To avoid any corresponding discussion on proper choices of further algorithmic parameters, we employ here a Poincar\'e section to each sample trajectory of the system at $y=0$, $\dot{y}<0$ (for different values of $b$) and construct OPTNs from $N = 10,000$ intersection points. The results are shown in Fig.~\ref{fig:bifurRossler} \textcolor{red}{and Supplementary Material Fig.~S5 for the same embedding dimension $D=6$, but two different values of the embedding delay $\tau$}. Based on these results, we conclude that the general behavior of the different SCMs and associated entropy measures closely resembles that reported for the logistic map in Fig.~\ref{fig:bifurcation}. Specifically, all measures trace the succession of bifurcations in the considered range of $b$. We outline more detailed follow-up investigations on the behavior of our OPTN based SCMs to such time-continuous systems directly as relevant topics for future work.

>>>>

The situation becomes even more challenging when our modified SCMs are considered, since they are based on transition frequencies instead of pattern frequencies. For $D!$ different ordinal patterns for embedding dimension $D$, there may exist $D!^2$ possible source--destination pairs. However, if we consider $\tau=1$ and transitions corresponding to shifting the time index by $\Delta t=1$ time step only, from the $D$ components of the source permutation $D-1$ also contribute to the corresponding destination pattern without changing their mutual order relationships \cite{BorgesAMC2019,PessaPRE2019}. Hence, in this special case the new state contributing to the destination pattern can only lead to at most $D$ different ordinal patterns, i.e., we have at most $D\cdot D!$ different pattern transitions. However, in other situations with $\tau>1$ and/or larger time shifts $\Delta t$ between successive ordinal patterns to be considered, up to $D!^2$ different pairs of patterns and, hence, transitions could occur. In general, we conclude that the far higher number of pattern transitions as compared to patterns themselves puts severe restrictions to the choice of $D$ in practical applications of the new SCMs in the context of time series of a given length $L$.
}

@@

\subsection{\textcolor{red}{Effect} of embedding parameters} \label{sec:embeddings}
There are two important algorithmic parameters in constructing OPTNs -- the embedding parameters $D$ and $\tau$. The choice of $D$ and $\tau$ can markedly affect the results for the associated SCMs since $D$ and $\tau$ should compromise the essential time scales of the dynamics in the underlying time series. In common applications of ordinal pattern analysis (e.g., the estimation of the permutation entropy $\mathcal{S}_O$), the choice of optimal (or just pragmatic) values of $D$ and $\tau$ depends on the field of application as systematically summarized in Ref.~\cite{Riedl2013}. {\color{red}Throughout the remainder of this work, we will consider $D$ as a free parameter, which is varied in the range of $D \in \{2,\ldots,7\}$, although for the sake of brevity, we will mainly show the results for $D = 6$ as illustration.

The associated choice of the embedding delay $\tau$ is commonly less crucial in nonlinear time series analysis methods. In our numerical examples discussed below, the corresponding selection will be mainly guided by the attempt to highlight the fact that lower entropy values are expected for periodic as compared to chaotic behavior. In order to cope with a single parameter only, we will particularly use $\tau=D$ for the artificial (deterministic and stochastic) model systems considered in the following, independent of the particular value of $D$. We emphasize that this choice is not based on specific considerations of performance optimization, but simply reflects convenience and serves solely illustrative purposes. Two examples for results obtained with $D=6$ and $\tau=1$ for two well-known model systems can be found in the Supplementary Material Sec.~SM2, Figs.~S4 and S5. In comparison with the corresponding results presented in the following, different values of $\tau$ seem to commonly lead to similar qualitative behaviors of our entropy and SCM characteristics, yet possibly different "resolution" (e.g., when inter-comparing the obtained numerical values for periodic and chaotic time series).}

@@

\textcolor{red}{More importantly, for sufficiently large $D$ the resulting complexity--entropy pairs} essentially cover exclusively the left uprising branch of the curve for the SCM \textcolor{red}{accounting for} transition frequencies (Figs.~\ref{fig:CElogistic}(b, e) and \ref{fig:CElogistic}(c, f)), which distinctively differs from the original permutation entropy based SCM. \textcolor{red}{In this regard, one may note that this branch implies an almost linear dependency between $\mathcal{C}_W$ and $\mathcal{H}_{W}$ (Fig.~\ref{fig:CElogistic}(e)), respectively, $\mathcal{C}_{E}$ and $\mathcal{H}_{E}$ (Fig.~\ref{fig:CElogistic}(f)), which especially applies for small entropy values within an increasingly wide range as $D$ increases. This linear increase of the maximum complexity versus entropy curve for sufficiently large embedding dimension is expected from theoretical considerations~\cite{SMCbook2010}. Some additional discussion on this observation can be found in the Supplementary Material Sec.~ SM3, supported by the corresponding Fig.~S6.}

@@

{\color{red}The behavior of the different SCMs relative to their respective lower and upper bounds can be further highlighted by considering normalized quantities. Specifically, for a given $\mathcal{H}$ and $D$, we define a normalized complexity 
\begin{equation}
\mathcal{C}^{N} = \frac{\mathcal{C} - \mathcal{C}_{min}}{\mathcal{C}_{max} - \mathcal{C}_{min}},
\end{equation}
\noindent
which is theoretically bound to the interval $[0,1]$. Large values of those normalized complexities close to 1 indicate a good alignment with the maximum complexity curve. The corresponding results for the logistic map are shown in Fig.~\ref{fig:CHlogCmax}. As expected, for the traditional SCM $\mathcal{C}_{O}$, the behavior is characterized by many points in the right half plane (large entropies, see Fig.~\ref{fig:CHlogCmax}(a)) for all considered embedding dimensions. In contrast, there are no points in the range of $\mathcal{H} >0.8$ for $\mathcal{C}_W$ and $\mathcal{C}_E$ (Fig.~\ref{fig:CHlogCmax}(b, c)). Note that in Fig.~\ref{fig:CHlogCmax}(c), some normalized complexity values larger than 1 arise at small entropies due to numerical artifacts when the control parameter $r$ is close to a bifurcation point, which results in long transients (as will be explained in Sec.~\ref{sec:transi}).}

@@

Especially for the new OPTN based SCMs, we find a partial overlap between the entropy values of fBm and the logistic map, while the associated SCM values are however clearly distinct except for the \textcolor{red}{fBm realizations} with the lowest entropy \textcolor{red}{values}, which correspond to the most nonstationary situations ($H$ close to 1), \textcolor{red}{which can be further seen in the Supplementary Material Fig.~S7. In general, we observe for all three complexity--entropy planes of fBm that with increasing $D$, the associated range of entropy values extends further towards the left (lower entropies). Taken both, upward and leftward shift of the complexity--entropy curves individually may therefore enhance the risk of confusing deterministic-chaotic and stochastic processes in the complexity--entropy plane even with our alternative SCMs. However, our results indicate that} combining information on the position in the complexity--entropy plane and on the stationarity could allow a clear distinction between chaotic and stochastic dynamics even in extreme situations \textcolor{red}{(e.g., a large value of $H$)}. Notably, for the stationary fGn, all \textcolor{red}{three complexity--entropy planes} are confined to very high entropy and low complexity values \textcolor{red}{(see also Supplementary Material Fig.~S8)} -- a range where no combinations have been found for the new transition frequency based SCMs for the logistic map.

@@

As a general observation, we find that all three SCMs clearly follow the changes in the bifurcation diagram, for instance, showing distinct values in periodic windows (highlighted by grey color in Fig.~\ref{fig:bifurcation} \textcolor{red}{and Supplementary Material Fig.~S4, where the obtained results are shown for the same embedding dimension $D=6$, but two different values of the embedding delay $\tau$). Notably, the distinction between periodic and chaotic windows in terms of all entropy and SCM values is more pronounced for $\tau=6$ as compared to $\tau=1$, which supports our choice of a higher embedding delay for obtaining the results discussed in the main text. However,} other than $\mathcal{C}_O$ the two transition frequency based SCMs behave in a similar way as the Lyapunov exponent. Specifically, as the control parameter $r$ is increased, the chaoticity level grows gradually, reaching a maximum at $r = 4$, which is reflected by an increase in statistical complexity. A similar increase is absent in the pattern frequency based complexity measure $\mathcal{C}_O$ (Fig.~\ref{fig:bifurcation}(c)). In this context, we note that it is an established fact that pattern frequency based SCMs often do not trace the growth in the chaoticity of the logistic map~\cite{MartinPLA2003}, which can however be improved by using Wooter's distance function instead of the Shannon-Jensen divergence employed in this work. From this point of view, the pattern transition frequency based SCMs $\mathcal{C}_W$ and $\mathcal{C}_E$ exhibit some more informative behavior in the sense that they track the growth of the level of chaoticity.

@@

While we have focused so far exclusively on the case of the time-discrete logistic map, it may be interesting to study whether a similar behavior can also be obtained for time-continuous deterministic dynamical systems. To this end, we will illustrate a corresponding analysis for the example of the chaotic R\"ossler system \cite{Roessler1976} 
\begin{eqnarray}
\dot{x} &=& -y-z, \nonumber \\
\dot{y} &=& x+0.2y, \\
\dot{z} &=& b+z(x-5.7) \nonumber
\end{eqnarray}
while varying the control parameter $b$. From a conceptual perspective, constructing OPTNs for time series from continuous systems faces certain additional practical challenges, including the proper choice of sampling frequency and embedding parameters, which depend on the particular time scales of the system. To avoid any corresponding discussion on proper choices of further algorithmic parameters, we employ here a Poincar\'e section to each sample trajectory of the system at $y=0$, $\dot{y}<0$ (for different values of $b$) and construct OPTNs from $N = 10,000$ intersection points. The results are shown in Fig.~\ref{fig:bifurRossler} \textcolor{red}{and Supplementary Material Fig.~S5 for the same embedding dimension $D=6$, but two different values of the embedding delay $\tau$. Again, we observe that the differences between periodic and chaotic windows become more pronounced for the larger value of the embedding delay $\tau$. In general, we find} that the \textcolor{red}{overall} behavior of the different SCMs and associated entropy measures closely resembles that reported for the logistic map in Fig.~\ref{fig:bifurcation}. Specifically, all measures trace the succession of bifurcations in the considered range of $b$. We outline more detailed follow-up investigations on the behavior of our OPTN based SCMs to such time-continuous systems directly as relevant topics for future work.
